Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-07-03 CGP Journal/main.tex
Moved semantic stuff, included Equation 1 in discussion

%TODO Consider adding 5-bit Parity or Equality, 11-bit Multiplexer.

@@

As a primary concept to our analysis, we determine the semantics of nodes
in the genome~\cite{mcphee:2008:treesemantics}.
For each input that is given to an individual, all of its nodes
will produce a single output.  These output values can be recorded and ordered.
As the domain in use is binary, the outputs can be viewed as a bit string, where
each bit indicates the value a node outputs when an individual is presented with a
specific input.  Similar recording can be done for other domains, but binary
allows for by far the simplest encoding.

\subsection{Never Active Nodes}
%TODO Somewhere in here you should talk about how Normal has the least number of active,
%which is probably related to how many never become active
In line with the suggestions made in~\cite{miller:2006:redundancy}, the evolved
number of active nodes across all problems and all algorithm configurations
in Table~\ref{tab:results}
was significantly lower than the genome size.  With the exception of a few combinations
with \emph{DAG} and \emph{Single}, the evolved genomes were over 50\% inactive.
\emph{Skip-Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
and 9\% of nodes in the genome active in the median runs of Parity, Encode, Decode,
and Multiply, respectively.  These final results fall in line with the predictions
made in~\cite{goldman:2013:ordering}, in that if the problems are ordered based on
their number of output locations, their evolved percentage of active nodes is also ordered.
%TODO Reference here again the probability function, and its ability to predict values

@@

The infrequently tried cause for a node never being active suggests that few
individuals are ever produced which have that node active, limiting the likelihood
of those individuals being selected regardless of potential fitness.  Under this
cause and combined with the theory given in~\cite{goldman:2013:ordering}, we
expect a strong positional bias in the location of the never active nodes;
we would expect nodes near the end of the genome to be the most likely to be never active.
Finally we would expect \emph{Reorder} and \emph{DAG} to have less or no positional bias.
All of these expectations match the results given in Table~\ref{tab:never_active}.

@@

@@

Generally, CGP appears to require the least number of evaluations when the genome
size far exceeds what is necessary for the problem at hand.  When properly tuned,
almost all variants had only a tiny fraction of the genome active in their final
solution.  Furthermore, a significant portion of these active nodes were found
to be redundant, removable using a very simple reduction algorithm.  The only
algorithm combinations that preferred more reasonable genome sizes were those using \emph{Single},
especially when paired with \emph{Reorder} or \emph{DAG}.  Yet we theorize this may
have more to do with \emph{Single} scaling poorly with the number of active nodes
than with it overcoming whatever causes traditional CGP to prefer enormous genomes.
While \emph{Reorder} and \emph{DAG} were able to use smaller genomes than \emph{Normal}
when paired with \emph{Single}, this decrease in size was not observed in any other
pairings.  As such, the suggestion in~\cite{goldman:2013:ordering} that these
techniques generally reduce genome size should be discarded.  Instead it is likely
more correct to say that \emph{Reorder} and \emph{DAG} are less detrimentally effected
by reducing the genome size.

>>>>

@@

\subsection{Never Active Nodes}
In line with the suggestions made in~\cite{miller:2006:redundancy}, the evolved
number of active nodes across all problems and all algorithm configurations
in Table~\ref{tab:results}
was significantly lower than the genome size.  With the exception of a few combinations
with \emph{DAG} and \emph{Single}, the evolved genomes were over 50\% inactive.
\emph{Skip-Normal}, our control representing normal CGP, had only 1\%, 3\%, 20\%,
and 9\% of nodes in the genome active in the median runs of Parity, Encode, Decode,
and Multiply, respectively.  These final results fall in line with the predictions
made in~\cite{goldman:2013:ordering}, in that if the problems are ordered based on
their number of output locations, their evolved percentage of active nodes is also ordered.
Furthermore, if Equation~\ref{eq:prob} is used to calculate the expected number of
active nodes for configurations using \emph{Normal} ordering, we find that in all
cases the evolved number is only marginally higher than the expectation.

@@

The infrequently tried cause for a node never being active suggests that few
individuals are ever produced which have that node active, limiting the likelihood
of those individuals being selected regardless of potential fitness.  Under this
cause and combined with Equation~\ref{eq:prob}, we
expect a strong positional bias in the location of the never active nodes;
we would expect nodes near the end of the genome to be the most likely to be never active.
Figure~\ref{fig:prob} shows the expected probability given by Equation~\ref{eq:prob},
using the same genome size, arity, number of inputs, and number of outputs as our
3-bit multiply.  This curve bears a striking resemblance to all three \emph{Normal}
plots in Table~\ref{tab:never_active}.
Finally, from~\cite{goldman:2013:ordering} we would expect \emph{Reorder} and \emph{DAG}
to have less or no positional bias, which also matches the results given in Table~\ref{tab:never_active}.

@@

In order to understand how each node is being used by CGP,
we first determine the semantics of each node~\cite{mcphee:2008:treesemantics}.
For each input that is given to an individual, all of its nodes
will produce a single output.  These output values can be recorded and ordered.
As the domain in use is binary, the outputs can be viewed as a bit string, where
each bit indicates the value a node outputs when an individual is presented with a
specific input.  Similar recording can be done for other domains, but binary
allows for by far the simplest encoding.

@@

Generally, CGP appears to require the least number of evaluations when the genome
size far exceeds what is necessary for the problem at hand.  When properly tuned,
almost all variants had only a tiny fraction of the genome active in their final
solution.  Furthermore, a significant portion of these active nodes were found
to be redundant; removable using a very simple reduction algorithm.  The only
algorithm combinations that preferred more reasonable genome sizes were those using \emph{Single},
especially when paired with \emph{Reorder} or \emph{DAG}.  Yet we theorize this may
have more to do with \emph{Single} scaling poorly with the number of active nodes
than with it overcoming whatever causes traditional CGP to prefer enormous genomes.
While \emph{Reorder} and \emph{DAG} were able to use smaller genomes than \emph{Normal}
when paired with \emph{Single}, this decrease in size was not observed in any other
pairings.  As such, the suggestion in~\cite{goldman:2013:ordering} that these
techniques generally reduce genome size should be discarded.  Instead it is likely
more correct to say that \emph{Reorder} and \emph{DAG} are less detrimentally effected
by reducing the genome size.
